### 2020-02-22 chapters/background.tex
HyperLogLog. PDIS delivery.

%1st paradigm, processing of large volumes of data
%oldest processing paradigm, simplest
Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user"}. Batch data sets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in the google search engine. For instance, grouping the results by keyword and counting how many times a keyword was searched for trend analysis purposes. Batch processing is the oldest paradigm between batch and stream processing. Historically speaking, Apache Hadoop \cite{hadoop} was one of the first widely known technologies for large-scale batch processing, created in 2005. Nearly a decade later, Apache Spark \cite{spark} was released. Spark is capable of doing not only on-disk processing like Hadoop but also in-memory processing, achieving lower processing times. As more and more data became available, the need for stream processing increased. Stream processing refers to the processing of data streams. A data stream is essentially an unbounded data set. This is in contrast to batch processing which works on a finite data set. This distinction is important since some computations are impossible to compute accurately on unbounded data sets. Such an example of an impossible computation on an infinite data set is grouping incoming data by an ID and counting distinct elements.
%On the other hand, stream processing refers to the processing of data streams. A data stream is essentially an unbounded data set. This is in contrast to batch processing which works on a finite data set. This distinction is important since some computations are impossible to compute accurately on unbounded data sets. Such an example of an impossible computation on an infinite data set is grouping incoming data by an ID and counting distinct elements. An example of such Stream Processing Engines (SPE) is Apache Flink \cite{flink}.

% background on batch systems, micro-batch, true streaming
%Batch processing is considered the oldest between both processing paradigms. Historically speaking, Apache Hadoop \cite{hadoop} was one of the first widely known technologies for large-scale batch processing, created in 2005. Nearly a decade later, Apache Spark \cite{spark} was released. Spark is capable of doing not only on-disk processing like Hadoop but also in-memory processing, achieving lower processing times. As more and more data became available, the more need for stream processing increased. % join this with previous batch intro
In order to provide a streaming solution, Apache Spark created the Spark Streaming module. Being a batch system in its core, Spark implements streaming as micro-batching. This approach handles streaming data by grouping incoming events into very small batches. Then, in order to process the entire defined window of data, it joins all micro-batches into a larger one and performs the desired computations. However, this creates an artificial barrier that does not really exist in the streaming context. More than conceptually different from a true stream processing system, processing a stream in a batch fashion, even if in small ones, presents some issues. First, since we are indeed working with batches, results will never be real-time updated per event. For example, if we use a batch size of 100 events, the system would only produce accurate results every 100 events. To handle system inactivity usually these micro-batch systems also have the notion of updating the aggregations with incomplete batches, if no event has entered the system after a pre-specified time period has elapsed. As such, micro-batching was created as a natural implementation of stream processing using the available batch processing systems of the time, but can not be considered true stream processing.

@@

An aggregation --- also known as a fold, reduce or accumulate --- is an operation that reduces a collection of values to a single one. Examples of such are the computation of data set sums, minima, averages or the number of distinct elements (cardinality). An aggregation may be exact or approximate. Exact aggregations compute a value from a sequence of elements, where no error or loss of resolution exists. On the other hand, approximate aggregations have an error margin to the associated aggregation value which is usually controllable by adjusting a set of parameters.

Aggregation is present in most programming languages and frameworks. Quoting from the JavaSE13 documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Hadoop's MapReduce paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one. \textit{Map} task will transform each item as desired and the \textit{reduce} task will merge the collection into a single value. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregated state. Doing this recursively for each item in the collection results in a single value: the aggregate value for the data set.

\subsection{Aggregation properties}

@@

These properties are used to evaluate SWAG algorithms. Different algorithms might only apply to certain families of aggregations. This way, the multiple SWAG algorithms presented throughout this Thesis will be bench-marked not only on their time and space complexity but also on the restrictions they impose regarding the aggregation' properties required.

@@

Unlike many other scenarios that focus on data at rest, streaming aggregations handle data in motion. While some aggregations can be computed on an infinite data set, often we want to compute these over a finite one. As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques. In doing so, we obtain a finite data set denoted as a sliding window and compute aggregations over it. The authors of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs). Once again let us consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs. How would we compute the average temperature per rack in the data center in the last 5 minutes? To better understand, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink \cite{flink}. Assuming all the temperatures can be queried from the \textit{Temperatures} table and that to each rack corresponds a different \textit{rack\_id} we have the following query:

@@

A sliding window aggregation (SWAG) algorithm allows the computation of an exact or approximate aggregation over a sliding window. A SWAG algorithm may restrict the set of possible to compute aggregations based on their properties. Different SWAG algorithms use different data structures for the window contents and the incremental aggregation state. Despite storing information using different structures, every SWAG algorithm must implement three methods:

\begin{itemize}
    \item \textit{insert} - inserts new value in the window
    \item \textit{evict} - evicts all values that no longer fit in the window
    \item \textit{query} - returns the aggregation state/value
\end{itemize}

>>>>

Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user"}. Batch data sets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in the google search engine. For instance, grouping the results by keyword and counting how many times a keyword was searched for trend analysis purposes. Batch processing is the oldest paradigm between batch and stream processing. Historically speaking, Apache Hadoop \cite{hadoop} was one of the first widely known technologies for large-scale batch processing, created in 2005. Nearly a decade later, Apache Spark \cite{spark} was released. Spark is capable of doing not only on-disk processing like Hadoop but also in-memory processing, achieving lower processing times. As more and more data became available, the need for stream processing increased. Stream processing refers to the processing of data streams. A data stream is essentially an unbounded data set. This is in contrast to batch processing which works on a finite data set. This distinction is important since some computations are impossible to compute accurately on unbounded data sets. Such an example of an impossible computation on an infinite data set is grouping incoming data by an ID and counting distinct elements.

In order to provide a streaming solution, Apache Spark created the Spark Streaming module. Being a batch system in its core, Spark implements streaming as micro-batching. This approach handles streaming data by grouping incoming events into very small batches. Then, in order to process the entire defined window of data, it joins all micro-batches into a larger one and performs the desired computations. However, this creates an artificial barrier that does not really exist in the streaming context. More than conceptually different from a true stream processing system, processing a stream in a batch fashion, even if in small ones, presents some issues. First, since we are indeed working with batches, results will never be real-time updated per event. For example, if we use a batch size of 100 events, the system would only produce accurate results every 100 events. To handle system inactivity usually these micro-batch systems also have the notion of updating the aggregations with incomplete batches, if no event has entered the system after a pre-specified time period has elapsed. As such, micro-batching was created as a natural implementation of stream processing using the available batch processing systems of the time, but can not be considered true stream processing.

@@

An aggregation --- also known as a fold, reduce or accumulate --- is an operation that reduces a collection of values to a single one. Examples of such are the computation of data set sums, minima, averages or the number of distinct elements (cardinality). An aggregation may be exact or approximate. Exact aggregations compute a value from a sequence of elements, where no error or loss of resolution exists. On the other hand, approximate aggregations have an error margin to the associated aggregation value which is usually controllable by adjusting a set of parameters. Approximate aggregations trade precision of results for efficient time and space usage. Such trade-off between accuracy and time/space efficiency will be made clear when discussing some of the existing approximate aggregators in Section \ref{sec:pds}.

Aggregations are present in most programming languages and frameworks. Quoting from the JavaSE13 documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Hadoop's MapReduce paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one. \textit{Map} tasks will transform each item as desired and the \textit{reduce} task will merge the collection into a single value. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregated state. Doing this recursively for each item in the collection results in a single value: the aggregate value for the data set.

\subsection{Aggregation properties}
\label{sec:agg-properties}

@@

%These properties are used to evaluate SWAG algorithms. Different algorithms might only apply to certain families of aggregations. This way, the multiple SWAG algorithms presented throughout this Thesis will be analyzed not only on their time and space complexity but also on the restrictions they impose regarding the aggregation' properties required.

@@

Unlike many other scenarios that focus on data at rest, streaming aggregations handle data in motion. While some aggregations can be computed on an infinite data set, often we want to compute these over a finite one. As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques. In doing so, we obtain a finite data set denoted as a sliding window and compute aggregations over it. The authors of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs). Once again let us consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs. How would we compute the average temperature per rack in the data center in the last 5 minutes? To better understand, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink \cite{flink}. Assuming all the temperatures can be queried from the \textit{Temperatures} table and that each rack corresponds a different \textit{rack\_id} we have the following query:

@@

A sliding window aggregation (SWAG) algorithm allows the computation of an exact or approximate aggregation over a sliding window. A SWAG algorithm may restrict the set of possible to compute aggregations based on their properties. Different SWAG algorithms use different data structures for the window contents and the incremental aggregation state.

The aggregation properties analyzed in Section \ref{sec:agg-properties} are used to evaluate SWAG algorithms. Different algorithms might only apply to certain families of aggregations. This way, the multiple SWAG algorithms presented throughout this Thesis will be analyzed not only on their time and space complexity but also on the restrictions they impose regarding the aggregation' properties required.

### 2020-02-22 chapters/sota.tex
HyperLogLog. PDIS delivery.

\subsection{A two window paradigm algorithm for change detection}
\label{subsec:2-window}

@@

This algorithm reduces change detection in data streams to testing whether two samples, the \textit{reference} and the \textit{sliding window} contents, are generated by different distributions. Consequently, the authors study the case of detecting differences in distribution between two samples. This is the purpose of the \textit{d} function. The authors evaluate many statistical tests as possible implementations of this \textit{d} function that must truly quantify an intuitive notion of change.

Experiments were made using the following statistical tests as the \textit{d} function: Wilcoxon signed-rank, $\phi$\textsubscript{A} and $\Xi$\textsubscript{A}. The Wilcoxon test is well known but the $\phi$\textsubscript{A} and the $\Xi$\textsubscript{A} statistical tests were defined by the authors. Since computing these three statistical tests for our windows contents takes identical time and space, we consider that a thorough analysis of each one is out of scope for this Thesis, as it does not provide relevant information.

@@

Initialization is not frequent. It occurs once at the beginning of execution and once per change detection upon which the algorithm resets the windows' contents. To provide an accurate time complexity, we take into account the probability \textit{P} of change occurring and being detected leading to a reset. Assuming \textit{P} is within $[0, 1]$, we have a time complexity of \textit{O(P(k + k log(2k)))} per reset --- \textit{i.e.} the probability of a reset occurring times \textit{O(k)} to initialize the windows' contents plus \textit{O(k log(2k))} to build the balanced tree.

@@

\section{Sliding Window Aggregation Algorithms} \label{sec:sota-swag-algs}
As discussed in Section \ref{sec:back-swag-algs} ...
Besides RFS and SOE other methods were proposed like DABA...
\subsection{DABA}
\subsubsection*{Time complexity analysis}
\subsubsection*{Space complexity analysis}
\subsubsection*{Applicability to our Hypothesis}

\section{Probabilistic Data Structures} \label{sec:pds}
In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs). However, if the aggregation we desire to compute can be approximated, there may exist more memory efficient methods. These aggregators are commonly denoted as sketches or probabilistic data structures and produce results orders-of magnitude faster then some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds. Each probabilistic data structure is used for efficient computation of one approximate aggregation, hence we say they are aggregation-specific.

@@

\subsection{Membership Queries and the Bloom filter}
% approx aggregation to compute
A Bloom filter \cite{BLOOM-BLOOMFILTER} approximate aggregator allows membership queries on a set of elements. Bloom filters answer the query: "Is element \textit{el} in the set of seen elements so far?". Being a probabilistic data structure and approximate aggregator, Bloom filters' answers have an error associated. \textit{False positive} matches are possible --- \textit{i.e.} saying \textit{el} is in the set when it is not --- but \textit{false negatives} are not --- \textit{i.e.} saying \textit{el} is not in the set but it actually is. In other words, a Bloom filter will accurately identify all items that do not belong in the set but will misclassify some items as being present in the set. Hence, a query returns either \textit{possibly in set} or \textit{definitely not in set}.

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=1.1]{figures/initial-bloom.png}
      \caption[Bloom filter bit array initial state]{Bloom filter bit array initial state of \textit{m}=7}
      \label{fig:initial-bloom-filter}
    \end{center}
\end{figure}

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/query-bloom.png}
      \caption[Bloom filter membership test]{Bloom filter membership test of element \textit{el}. Finding at least one '0' indicates that it is not in the set.}
      \label{fig:bloom-filter}
    \end{center}
\end{figure}

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/fp-bloom-1.png}
      \caption[Bloom filter insertion]{Insertion of \textit{el1} in the Bloom filter}
      \label{fig:bloom-filter-fp1}
    \end{center}
\end{figure}

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/fp-bloom-2.png}
      \caption[Bloom filter insertion]{Insertion of \textit{el2} in the Bloom filter}
      \label{fig:bloom-filter-fp2}
    \end{center}
\end{figure}

@@

% FOR EACH SKETCH DO:
% approx aggregation to compute
% data structures
% algorithm for insertion
% algorithm for query
% error bound related with parameters
% time complexity analysis
% space complexity analysis
% Applicability to our Hypothesis
\subsection{Item Frequency and the Count-Min Sketch}
% approx aggregation to compute
A Count–Min Sketch (CMS) is an approximate aggregator that estimates the frequency of each element in the data set. It is named after the two basic operations used to provide the estimate: counting (\textit{count}) and computing the minimum (\textit{min}). The CMS is first introduced in \cite{Cormode-CMS}.

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.5]{figures/initial-cms.png}
      \caption[Count-Min Sketch matrix initial state]{Count-Min Sketch matrix initial state of \textit{d}=4 and \textit{w}=9}
      \label{fig:initial-cms}
    \end{center}
\end{figure}

@@

Figure \ref{fig:query-cms} illustrates this procedure.

@@

%The CMS is not always correct because...
\subsubsection*{Time complexity analysis}
When adding an element, the Count-Min Sketch loops through each row and applies a constant in time hash function, updating the value of the target cell. Hence, insertion loops through \textit{d} rows and has time complexity of \textit{O(d)}. Similarly, querying the frequency of an element \textit{el} consists in computing the minimum of all row values. Since both insertion and query have constant time complexity, the CMS is constant in time.

@@

\subsection{Cardinality Estimation and the HyperLogLog}
In the paper \cite{Flajolet-PCA}, a class of probabilistic counting algorithms is introduced to estimate the number of distinct elements based on \textit{"bit pattern observables"} in the binary representation of the hashed values --- \textit{i.e.} patterns related to the number and positions of 0's and 1's in a binary string.

\subsubsection{LogLog}
Later on, the same authors propose the LogLog algorithm \cite{Flajolet-LogLog}. In LogLog, the \textit{bit pattern observable} recorded from each item's hash value is the position of the leftmost 1-bit. The authors claim it is related to the total number of distinct elements in the data set.

\subsubsection{HyperLogLog}
A couple of years later, the same authors propose HyperLogLog (HLL) in \cite{Flajolet-HLL}. HLL solves the problem of counting distinct elements in a data set, also known as the cardinality of the data set \cite{Flajolet-HLL}. HLL gives an approximation of the cardinality of the data set.

In HyperLogLog, an item is hashed into a binary string. Based on the first \textit{n} bits, the string is passed on to one of the \textit{$2^n$ buckets}. After discarding the first \textit{n} bits, the number of trailing 0's plus 1 is counted. In other words, the position of the leftmost 1-bit is recorded, like in LogLog. This is what is stored in each bucket: the maximum value of the position of the leftmost 1-bit of all seen items so far.

%TODO \textbf{HYPERLOGLOG HOW IT WORKS IMAGE}

Finally, the estimate for the number of distinct elements will be $2^p$, where \textit{p} is the harmonic mean of all bucket values holding the maximum leftmost 1-bit position value. This computation of a harmonic mean over \textit{m} buckets is called stochastic averaging and is the main difference between HyperLogLog and LogLog. Given the nature of streaming systems, it is also important to notice that this algorithm is suitable for a distributed system since each of the buckets can be updated independently.

\subsubsection*{Time complexity analysis}
Adding an item is tantamount to hashing it, determining the corresponding bucket based on the first \textit{n} bits, count the leftmost 1-bit position and update the bucket value if need be. Estimating the cardinality of the set amounts to finding the maximum value \textit{p} between all buckets and computing $2^p$. Hence, both inserting a new element and estimating the cardinality take constant time --- \textit{O(1)}.

\subsubsection*{Space complexity analysis}
Quoting Flajolet et al. about the memory consumption and error in the approximations obtained in their experiments in  \cite{Flajolet-HLL}: 
\say{cardinalities till values over N = $10^9$ can be estimated with a typical accuracy of 2\% using 1.5kB (kilobyte) of storage.}

HyperLogLog uses \textit{m} buckets that store a single integer value. This gives it complexity of \textit{O(m)} where \textit{m} is a small constant. Hence, HyperLogLog has constant space complexity --- \textit{O(1)}.

\subsubsection*{Applicability to our Hypothesis}
Similar to the previously studied probabilistic data structures, HyperLogLog (HLL) has constant space and time complexity while giving an approximation of a set's cardinality. Hence, HLLs are considered valuable methods for our work.

>>>>

\subsection*{A two window paradigm algorithm for change detection}
\label{subsec:2-window}

@@

This algorithm reduces change detection in data streams to testing whether two samples, the \textit{reference} and the \textit{sliding window} contents, are generated by different distributions. Consequently, the authors study the case of detecting differences in distribution between two samples. This is the purpose of the \textit{d} function. The authors evaluate several statistical methods as possible implementations of this \textit{d} function in order to encode an intuitive notion of change.

The following statistical tests for the \textit{d} function were used: Wilcoxon signed-rank, $\phi$\textsubscript{A} and $\Xi$\textsubscript{A}. The Wilcoxon test is well known but the $\phi$\textsubscript{A} and the $\Xi$\textsubscript{A} statistical tests were defined by the authors. Since computing these three statistical tests for our windows contents takes identical time and space, we consider that a thorough analysis of each one is out of scope for this Thesis, as it does not provide relevant information.

@@

Initialization is not frequent. It occurs once at the beginning of execution and once per change detection upon which the algorithm resets the windows' contents. To provide an accurate time complexity, we define \textit{P} as the probability of a change occurring and being detected leading to a reset. Assuming \textit{P} is within $[0, 1]$, we have a time complexity of \textit{O(P(k + k log(2k)))} per reset --- \textit{i.e.} the probability of a reset occurring times \textit{O(k)} to initialize the windows' contents plus \textit{O(k log(2k))} to build the balanced tree.

@@

%\section{Sliding Window Aggregation Algorithms} \label{sec:sota-swag-algs}
%As discussed in Section \ref{sec:back-swag-algs} ...
%Besides RFS and SOE other methods were proposed like DABA...
%\subsection{DABA}
%\subsubsection*{Time complexity analysis}
%\subsubsection*{Space complexity analysis}
%\subsubsection*{Applicability to our Hypothesis}

\section{Probabilistic Data Structures} \label{sec:pds}
%TODO: UNCOMMENT
%In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs). However, if the aggregation we desire to compute can be approximated, there may exist more memory efficient methods. These aggregators are commonly denoted as sketches or probabilistic data structures and produce results orders-of magnitude faster then some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds. Each probabilistic data structure is used for efficient computation of one approximate aggregation, hence we say they are aggregation-specific.

@@

These aggregators are commonly denoted as sketches and produce results orders-of magnitude faster then some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds. Each probabilistic data structure is used for efficient computation of one approximate aggregation, hence we say they are aggregation-specific.

\subsection{Membership Queries and the Bloom filter} \label{sec:bloom}
% approx aggregation to compute
A Bloom filter \cite{BLOOM-BLOOMFILTER} approximate aggregator allows membership queries on a set of elements. Bloom filters answer the query: "Is element \textit{el} in the set of seen elements so far?". Being a probabilistic data structure and approximate aggregator, Bloom filters' answers have an error associated. \textit{False positive} matches are possible --- \textit{i.e.} saying \textit{el} is in the set when it is not --- but \textit{false negatives} are not --- \textit{i.e.} saying \textit{el} is not in the set but it actually is. In other words, a Bloom filter will accurately identify all items that do not belong in the set but will misclassify some items as being present in the set. Hence, a query returns either \textit{possibly in set} or \textit{definitely not in set}.

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=1.1]{figures/initial-bloom.png}
      \caption[Bloom filter initial state]{Bloom filter bit array initial state of \textit{m}=7}
      \label{fig:initial-bloom-filter}
    \end{center}
\end{figure}

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/query-bloom.png}
      \caption[Bloom filter membership query]{Bloom filter membership test of element \textit{el}. Finding at least one '0' indicates that it is not in the set.}
      \label{fig:bloom-filter}
    \end{center}
\end{figure}

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/fp-bloom-1.png}
      \caption[Bloom filter insertion of \textit{el1}]{Insertion of \textit{el1} in the Bloom filter}
      \label{fig:bloom-filter-fp1}
    \end{center}
\end{figure}

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/fp-bloom-2.png}
      \caption[Bloom filter insertion of \textit{el2}]{Insertion of \textit{el2} in the Bloom filter}
      \label{fig:bloom-filter-fp2}
    \end{center}
\end{figure}

@@

\subsection{Item Frequency and the Count-Min Sketch}
% approx aggregation to compute
The Count-Min Sketch (CMS) uses two principles from the discussed Bloom filters in Section \ref{sec:bloom}. First, Bloom filters show that precision can be sacrificed to achieve space savings. Second, they are very similar at a technical level.

A Count–Min Sketch \cite{Cormode-CMS} is an approximate aggregator that estimates the frequency of each element in the data set. It is named after the two basic operations used to provide the estimate: counting (\textit{count}) and computing the minimum (\textit{min}).

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.5]{figures/initial-cms.png}
      \caption[Count-Min Sketch initial state]{Count-Min Sketch matrix initial state of \textit{d}=4 and \textit{w}=9}
      \label{fig:initial-cms}
    \end{center}
\end{figure}

@@

Figure \ref{fig:query-cms} illustrates this procedure, where the result of the query "What is the frequency of \textit{el}?" returns "2".

@@

% error bound related with parameters
\subsubsection*{Optimal number of hash functions} 
For an element \textit{el}, each row will have one position incremented by \textit{el}'s true frequency. However, multiple elements may be mapped to the same position within a row, thus overlapping their frequencies. Hence, for an element \textit{el}, the CMS provides a frequency estimate that is greater or equal to the true frequency of \textit{el} in the set.

The number of hash functions \textit{d} to be used depends on the desired error probability. To achieve an error probability $\delta$, \textit{d} is given by the following expression:
\begin{equation}
    d = \lceil ln \frac{1}{\delta} \rceil
\end{equation}
For example, to achieve an error $\delta$ = 1\%, \textit{d} = 5 hash functions are necessary.

\subsubsection*{Time complexity analysis}
When adding an element, the Count-Min Sketch loops through each row and applies a constant in time hash function, updating the value of the target cell. Hence, insertion loops through \textit{d} rows and has time complexity of \textit{O(d)}. Similarly, querying the frequency of an element \textit{el} consists in computing the minimum of all row values. Since both insertion and query have constant time complexity, the CMS is constant in time.

@@

\subsection{Cardinality Estimation and the HyperLogLog}
A set of probabilistic counting algorithms is introduced by Flajolet et al. in \cite{Flajolet-PCA} to estimate the number of distinct elements of a set --- also known as the cardinality of the set. These techniques are based on \textit{"bit-pattern observables"} in the binary representation of the hashed values. Bit-pattern observables are defined by the authors as \textit{"patterns of bits occurring at the beginning of the (binary) S-values"}, with \textit{S} as the target set.

Later on, the authors propose the LogLog algorithm \cite{Flajolet-LogLog}. In LogLog, the bit-pattern observable recorded for each item is the position of the leftmost 1-bit, denoted as \textit{P}. The authors claim it is related to the total number of distinct elements in the data set in that it is \textit{"more or less a likely indication that the cardinality of S is at least $2^P$"}.

% FOR EACH SKETCH DO:
% approx aggregation to compute
% data structures
% algorithm for insertion
% algorithm for query
% error bound related with parameters
% time complexity analysis
% space complexity analysis
% Applicability to our Hypothesis
\subsubsection*{HyperLogLog}
A couple of years later, the same authors propose HyperLogLog (HLL) \cite{Flajolet-HLL}, an approximate aggregator to compute the cardinality of a set.

HLL uses the same bit-pattern observable as LogLog --- \textit{P}, the position of the leftmost 1-bit. With LogLog, the authors realize that storing only one observation produces extremely inaccurate predictions due to the high variability of a single value. The proposed solution is to apply \textit{stochastic averaging}. As the authors state, stochastic averaging \textit{"consists in emulating the effect of m experiments"}. Effectively speaking, the authors propose dividing the input stream into \textit{m} substreams or buckets, maintaining one observable per bucket and computing \textit{P} as an average of all bucket values (stochastic averaging).

\subsubsection*{Data Structure}
HyperLogLog (HLL) maintains \textit{m} bit-pattern observables, each being the leftmost 1-bit position, \textit{P}. Since each position is a single integer, HLL maintains \textit{m} integer counters, as seen in Figure \ref{fig:hll-ds}. Each of these counters holds the maximum value for the bit-pattern observable seen so far. The first bit is assumed to be in position "1" and all counters begin with "0".

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.6]{figures/hll-ds.png}
      \caption[HyperLogLog initial state]{HyperLogLog with \textit{m}=4 observable buckets}
      \label{fig:hll-ds}
    \end{center}
\end{figure}

\subsubsection*{Element Insertion}
In HyperLogLog, new elements are hashed into a binary string. The first \textit{n} bits of the hashed representation will map to one of the \textit{m}=$2^n$ buckets. Effectively speaking, the first \textit{n} bits work as a \textit{bucket ID}. The bit-pattern observable \textit{P} --- the position of the leftmost 1-bit --- is counted from the binary string discarding the first \textit{n} bits. The observable \textit{P} is stored in the pre-determined bucket if it is greater than the value the bucket contains.

Figure \ref{fig:hll-insert} demonstrates such procedure for a new element \textit{el}. The hash representation of \textit{el}, using \textit{n}=2 bits, maps to bucket \textit{10}. Discarding the first \textit{n}=2 bits and counting from 1, the position of the leftmost 1-bit is \textit{P}=4. The previous value of bucket \textit{10} was 0 and because \textit{P}=4 is greater than 0 the bucket value is updated.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=1]{figures/hll-insertion.png}
      \caption[HyperLogLog insertion]{HyperLogLog \textit{el} insertion with \textit{m}=4 and \textit{n}=2 bits}
      \label{fig:hll-insert}
    \end{center}
\end{figure}

\subsubsection*{Cardinality Query}
The cardinality of the set is estimated to be $2^\textit{P}$ where the bit-pattern observable \textit{P} is the position of the leftmost 1-bit. As discussed, HyperLogLog (HLL) differentiates from LogLog in that it applies a stochastic averaging process. Hence, the value of \textit{P} will be the harmonic mean of all \textit{m} buckets.

\subsubsection*{Time complexity analysis}
Adding an item is tantamount to hashing it, determining the corresponding bucket based on the first \textit{n} bits, counting the leftmost 1-bit position and updating the bucket value if need be. Estimating the cardinality of the set amounts to finding the maximum value \textit{P} between all buckets and computing $2^P$. Hence, both inserting a new element and estimating the cardinality of the set take constant time.

\subsubsection*{Space complexity analysis}
Quoting Flajolet et al. about the memory consumption and error in the approximations obtained in their experiments in the original HyperLogLog paper \cite{Flajolet-HLL}: 
\say{cardinalities till values over N = $10^9$ can be estimated with a typical accuracy of 2\% using 1.5kB (kilobyte) of storage.}

HyperLogLog uses \textit{m} buckets that store a single integer value, thus the space complexity is \textit{O(m)} where \textit{m} is a rather small constant. Hence, HyperLogLog has constant space complexity.

\subsubsection*{Applicability to our Hypothesis}
Similar to the previously studied approximate aggregators, HyperLogLog (HLL) has constant space and time complexity and provides an estimate of a set's cardinality. Hence, HLLs are considered valuable aggregators for our Thesis.
